Multivariate Neuroanalysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests

### 2013-11-06 paper.tex
ENH:  Getting closer.

%\documentclass[preprint,authoryear,review,12pt]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label2]{Jeffrey T.~Duda}
\author[label1]{Christopher R.~Durst}
\author[label1]{Max Wintermark}
\author[label1]{James C.~Gee}
\author[label1]{Murray C.~Grossman}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, University of Pennsylvania,
                Philadelphia, PA}

@@

\linenumbers

@@

\end{frontmatter}
%
%
\newpage

@@

In addition to describing ANTsR basics and how it can be
used generally in multivariate neuroimaging studies, we showcase its use in
the particularly difficult application of supervised brain tumor
segmentation \cite{bauer2013} for which we use the widely popular 
random forests machine learning technique.  Although random forests have
been proposed previously in the literature for, among other things, 
supervised brain segmentation (e.g. \cite{geremia2011}), 
we demonstrate that ANTsR significantly facilitates the construction 
of a fully functional, parallelizable workflow for such a task and 
that the performance of our proposed framework has proven extremely in 
actual practice, specifically
 in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
Challenge (see Figure \ref{fig:award}).%
\footnote{
http://martinos.org/qtim/miccai2013/
}
\begin{figure}
  \centering
  \includegraphics[width=80mm]{Figures/award.pdf}
  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
  \label{fig:award}
\end{figure}

@@

The ANTsR package is publicly available on the github project hosting service.%
\footnote{
https://github.com/stnava/ANTsR
}
Prior to installation of ANTsR, several external R packages
need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
\verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
\verb#rgl#, and \verb#misc3d#.  Additionally, in order
to perform the supervised brain segmentation as described 
in later sections, one needs to also install the packages
\verb#randomForest#, \verb#snowfall#, \verb#rlecuyer#,
and \verb#ggplot2#.%
\footnote{
Packages are easily installed using the {\tt install.packages()} R function.
}

@@

@@

Supervised techniques generally consist of a training phase
for model construction followed by prediction using the previously
generated model.  For supervised brain tumor segmentation, 
a set of training data consisting of labeled brain images 
(e.g. csf, gray matter, white matter, 
necrosis, edema, enhancing and  non-enhancing tumor) is
used to construct a predictive model.  Although other 
classification techniques have been used to segment
brain tumors (e.g. support vector machines \citep{bauer2011}),
random forest models have proven particularly successful.

In subsequent sections, we detail the generic
random forest modeling framework which takes as input
a set of feature images (also described in a later section)
and labels for each training subject 
and outputs a statistical model.  This model, in turn, can then
be used to provide a probabilistic estimate of the labels in an
unlabeled subject.

%One of the core extensions that we provide in this work is
%the use of concatenated random forests for improved probabilistic estimation
%of the labels.
%As we demonstrate in the Results section, the set of feature
%images employed work sufficiently well for good performance
%on the training data.  However, we discovered
%that further improvements could be gained by using the probabilistic 
%label estimates as input to enhance preprocessing before a 
%second round of feature image generation including 
%modality-specific, prior-based $n$-tissue segmentation and 
%random forest model creation.

Several machine learning concepts were integrated to create 
the random forests framework first articulated in its entirety by Breiman
et al. \cite{breiman2001} for performing classification/regression.  
Although decision trees had been previously explored in the literature, 
it was the success of ``boosting''-style machine learning 
techniques, such as AdaBoost \cite{schapire1990,freund1997}, which influenced 
the aggregation of such decision trees into ``forests'' 
with randomized node optimization for improved
classification/regression performance \cite{ho1995,amit1997}.
The final element of bootstrap aggregating or ``bagging'' (i.e.
random sampling of the training data) was
introduced by Breiman \cite{breiman1996} to achieve improved
accuracy.

@@

One of the principal advantages of R is the extensive community of
developers  who have contributed on the order of thousands of packages 
extending R's capabilities beyond its core functionality.
Most relevant for the work described in this paper
is the \verb#randomForest# package developed from Breiman's original
Fortran code by Liaw and Wiener \citep{liaw2002}.

@@

Training data consisted of multi-modal brain MRI (T1, T2, FLAIR, and 
post-Gadolinium T1) from 30 glioma patients (both low, $n=10$, and high-grade, $n=20$,
and with and without resection).  For each subject, the T1, T2, and 
FLAIR MRI were linearly registered to the post-contrast T1.  Subsequently,
the brains were skull-stripped and resampled to 1 mm isotropic resolution.
Testing data was processed similarly and released during the course of the
challenge in two sets denoted as ``LeaderBoard'' and ``Challenge'' data.  
The former consisted of 21 and 4 high and low-grade tumor patients, respectively,
whereas the latter comprised 10 high-grade only patients.

@@

In terms of implementation, this algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the Kirby multivariate symmetrical template consisting of FA, FLAIR,  
T1, and T2 components.  Although DTI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep{price2003} to warrant investigation in our future work.

@@

\begin{figure*}
  \centering
  \includegraphics[width=180mm]{Figures/featureImages.pdf}
  \caption{Representative feature images derived from the BRATS\_HG0301 ``challenge'' data set.
           Neighborhood statistical images for each modality were generated by calculating a 
           given statistic within
           a specified neighborhood radius.  Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific difference images with the
           corresponding symmetric template itself and with respect to the contralateral side.   
           This mapping is also used to produce the log Jacobian image.  Finally, the T1 - TC image
           is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
\end{figure*}

The workflow for estimating tumor-based labeling from multi-modal 
images involves the following steps:
\begin{enumerate}
  \item Symmetric multivariate template construction \cite{avants2010}
 using the data described in \cite{landman2011}.
  \item Image preprocessing:
    \begin{itemize}
      \item Windowing intensities (quantiles $[0.01,0.99]$).
      \item N4 bias correction \cite{tustison2010}.
      \item Rescaling intensity range to $[0,1]$.  
    \end{itemize}
  \item Stage 1 (GMM) processing:
  \begin{itemize}
    \item generation of feature images,
    \item construction of the Stage 1 RF model and probability images.
  \end{itemize}
  \item Stage 2 processing:
  \begin{itemize}
    \item generation of single-modality MAP-MRF images using the Stage 1 RF probability images as spatial priors,
    \item construction of the Stage 2 RF model and labelings.
  \end{itemize}
  \item Refinement of Stage 2 labelings using a heuristically-derived binary morphological processing protocol.
\end{enumerate}

Key to any supervised regression or classification protocol are the 
selected features for training and subsequent prediction.  Based on previous
work and our own experience, the following feature images were selected:
\begin{itemize}
  \item Per modality (FLAIR, T1, T1C, T2)
    \begin{itemize}
      \item First-order neighborhood statistical images:
            mean, variance, skewness, and entropy. 
            Neighborhood radius $\in \{1,3\}$.
    \item GMM (stage 1) and MAP-MRF (stage 2) posteriors: CSF, gray matter, white 
          matter, necrosis, edema, non-enhancing tumor and enhancing tumor (or a
          subset for the simulated data).
    \item GMM (stage 1) and MAP-MRF (stage 2) connected component geometry 
          features:  distance to tumor core label, volume, volume to surface area ratio, eccentricity, and elongation
    \item Template-based:  symmetric template difference and 
          contralateral difference with Gaussian smoothing ($\sigma = 4$mm).
    \end{itemize}
  \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
    log Jacobian image, and  (T1C - T1) difference image.
\end{itemize}

For each modality, we create four first-order statistical feature images,
five Gaussian mixture model (GMM)-based posterior probability feature images,
four geometry features generated from the GMM posterior probability images
based on connected components, and two difference images using symmetric template
construction for a total of 4 modalities $\times$ (4 + 5 + 4 + 2) feature images per modality $=$ 60 total
feature images.  We employ two additional images consisting of the 
Euclidean distance image \cite{maurer2003} created from the skull-stripped 
binary mask rescaled to the range $[0,1]$ and the log
Jacobian image derived from the spatial normalization of the symmetric multivariate template and individual subject images.  Given the intensity corrected images
the corresponding multivariate template images, and a brain mask for each subject
creation of all feature images is performed using the script \verb#createFeatureImages.sh#.

Prior  cluster centers for specific tissue types learned from training data \cite{reynolds2009} are used in the GMM to create multiple feature images.  
Given $M$ tissue types (e.g. CSF, gray matter,
white matter, edema, and tumor), a GMM formulates the 
probability distribution at each voxel, $\mathbf{x}$, as the
sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
\begin{align}
p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
\end{align}
where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
determining the parameters of the GMM is maximum likelihood 
estimation which can performed using the Atropos segmentation 
tool \cite{avants2011}.  In contrast to previous generative
modeling approaches for multi-modal tumor segmentation 
(e.g. \cite{prastawa2003}), we do not use multivariate 
Gaussians to specify tissue probabilities but rather incorporate each
univariate probability map into the feature vector of the training
data.  As pointed out in \cite{menze2010}, multivariate modeling
might obscure the distinct biological information provided by each 
modality.  Instead, we let the random forest construction 
process determine the optimal combination of such multivariate
information.
Additionally, maximum posterior labeling from the GMM processing
is used to determine the connected components for each label.  
Geometric features (assigned voxel-wise) include the physical volumes 
of each connected component including the volume to surface area ratio, 
the elongation, and eccentricity.

@@

\subsubsection{Prediction:  Applying Random Forest Models}

\begin{figure}
  \centering
  \begin{tabular}{c}
    \includegraphics[width=85mm]{Figures/pipeline.pdf}
  \end{tabular}
  \caption{Diagrammatic workflow for the proposed RF model prediction.  The feature
  images are first generated from the input MRI and symmetric multivariate template. 
  After organization of the feature images in a csv file, the stage 1 RF model is 
  applied which results in a first stage classification voting output.  This output
  is used to generate an additional set of feature images for the second classification
  stage.  The final refinement process entails a series of binary morphological 
  operations which results in the final labeled brain.  In the diagram we also list
  the files used to perform each set of functions which are available as open source.
  }
  \label{fig:pipeline}
\end{figure}

\begin{figure*}
  \centering
  \begin{tabular}{cc}
  \includegraphics[width=80mm]{Figures/BRATS_HG_GMM.pdf} &
  \includegraphics[width=80mm]{Figures/BRATS_HG_MAP_MRF.pdf} \\
  \includegraphics[width=80mm]{Figures/BRATS_LG_GMM.pdf} &
  \includegraphics[width=80mm]{Figures/BRATS_LG_MAP_MRF.pdf}
  \end{tabular}
  \caption{Feature images from the BRATS\_HG0004 data set.}
\end{figure*}

@@

\begin{table*}
\caption{Dice scores from the MICCAI 2012 BRATs Study}
\begin{center}
\begin{tabular*}{0.975\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c}
\toprule
{} & \multicolumn{2}{c}{High-grade (real)} & \multicolumn{2}{c}{Low-grade (real)} & \multicolumn{2}{c}{High-grade (simulated)} & \multicolumn{2}{c}{Low-grade (simulated)}\\
{\bf Method} & Edema & Tumor & Edema & Tumor & Edema & Tumor & Edema & Tumor\\
\midrule
\cite{zikic2012} & {$0.70 \pm 0.09$} & {$0.71 \pm 0.24$} & {$0.44 \pm 0.18$} & {$0.62 \pm 0.27$} & {$0.65 \pm 0.27$} & {$0.90 \pm 0.05$} & {$0.55 \pm 0.23$} & {$0.71 \pm 0.20$} \\
\cite{bauer2012} & {$0.61 \pm 0.15$} & {$0.62 \pm 0.27$} & {$0.35 \pm 0.18$} & {$0.49 \pm 0.26$} & {$0.68 \pm 0.26$} & {$0.90 \pm 0.06$} & {$0.57 \pm 0.24$} & {$0.74 \pm 0.10$} \\
ANTsR & {$0.65 \pm 0.15$} & {$0.66 \pm 0.28$} & {$0.49 \pm 0.16$} & {$0.65 \pm 0.21$} & {$0.68 \pm 0.25$} & {$0.91 \pm 0.08$} & {$0.61 \pm 0.25$} & {$0.84 \pm 0.09$} \\
w/ Atropos & {$0.68 \pm 0.15$} & {$0.67 \pm 0.30$} & {$0.50 \pm 0.15$} & {$0.67 \pm 0.23$} & {$0.74 \pm 0.26$} & {$0.92 \pm 0.09$} & {$0.65 \pm 0.26$} & {$0.84\pm 0.08$} \\
\bottomrule
\end{tabular*}
\end{center}
\end{table*}

>>>>

% \documentclass[preprint,authoryear,review,12pt]{elsarticle}
\documentclass[final,5p,times,twocolumn]{elsarticle}

@@

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label1]{Max Wintermark}
\author[label1]{Christopher R.~Durst}
\author[label2]{Jeffrey T.~Duda}
\author[label1]{James C.~Gee}
\author[label3]{Murray C.~Grossman}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, 
                 Department of Radiology, University of Pennsylvania,
                Philadelphia, PA}
\address[label3]{Department of Neurology, University of Pennsylvania,
                Philadelphia, PA}

@@

@@

\end{frontmatter}
%\linenumbers

%
%
\newpage

@@

In addition to describing ANTsR basics and how it can be
used generally in multivariate neuroimaging studies, we showcase its use in
the particularly difficult application of supervised brain tumor
segmentation \cite{bauer2013} for which we use the widely popular 
random forests machine learning technique.  Although random forests have
been proposed previously in the literature for, among other things, 
supervised brain segmentation (e.g. \cite{geremia2011}), 
we demonstrate that ANTsR significantly facilitates the construction 
of a fully functional, parallelizable workflow for such a task and 
that the performance of our proposed framework has proven extremely in 
actual practice, specifically
 in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
Challenge (see Figure \ref{fig:award}).%
\footnote{
http://martinos.org/qtim/miccai2013/
}
\begin{figure}
  \centering
  \includegraphics[width=90mm]{Figures/award.pdf}
  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
  \label{fig:award}
\end{figure}

@@

The ANTsR package is publicly available on the github project hosting service.%
\footnote{
https://github.com/stnava/ANTsR
}
Prior to installation of ANTsR, several external R packages
need to be installed including: \verb#Rcpp#, \verb#signal#, \verb#timeSeries#, 
\verb#mFilter#, \verb#doParallel#, \verb#robust#, \verb#magic#, \verb#knitr#, \verb#pixmap#, 
\verb#rgl#, and \verb#misc3d#.  Additionally, in order
to perform the supervised brain segmentation as described 
in later sections, one needs to also install the packages
\verb#randomForest#, \verb#snowfall#, \verb#rlecuyer#,
and \verb#ggplot2#.%
\footnote{
Packages are easily installed using the {\tt install.packages()} R mechanism.
}

@@

As mentioned earlier, we have made this entire framework
available as open source.  In addition to the ANTsR repository
already on github which houses both ANTs and ANTsR functionality, 
we created a special github repository specifically for this work
containing figures, references, and text.%
\footnote{
https://github.com/ntustison/ANTsAndArboles
}
Also, we posted all
scripts (R, shell, and perl) used to coordinate the ANTsR processing 
including (along with a brief description of each script):
\begin{itemize}
  \item {\tt applyModel.R}:  applies a random forest model to a new 
  feature data set from a testing subject resulting in a set of probability
  images (one for each label).
  \item {\tt applyTumorSegmentationModel.sh}:  generates the new feature image set 
  from the testing MRI (by calling {\tt createFeatureImages.sh}).
  organizes the file names in a csv file (via {\tt createCSVFileFromModel.R}),
  and applies the random forest model using {\tt applyModel.R}. 
  \item {\tt applyTumorSegmentationModelForCohort.pl}:  Coordinates tumor 
  segmentation on the computational cluster for a given cohort.
  \item {\tt createCSVFileFromModel.R}:  organizes the set of feature image
  file names in a csv file for input into {\tt applyModel.R}.
  \item {\tt createFeatureImages.sh};  creates the set of feature images given
  a set of co-registered input MRI from a single subject. 
  \item {\tt createModel.R}:  creates random forest model given the input csv 
  file of the feature image file names for all training data and set of truth
  label maps.
  \item {\tt plotVariableImportance.R}:  produces a plot of the importance 
  (`{\tt MeanDecreaseAccuracy}') of each feature variable used 
  in constructing the model.
\end{itemize}
We also include a fully functional 2-D example which performs
both testing and training on sample challenge data.

@@

Supervised techniques generally consist of a training phase
for model construction followed by prediction using the 
generated model.  For supervised brain tumor segmentation, 
a set of training data consisting of labeled brain images 
(e.g. csf, gray matter, white matter, 
necrosis, edema, enhancing and  non-enhancing tumor) is
used to construct a predictive model.  Although other 
classification techniques have been used to segment
brain tumors (e.g. support vector machines \citep{bauer2011}),
random forest models have proven particularly successful.

Several machine learning concepts were integrated to create 
the random forests framework first articulated in its entirety by Breiman
et al. \cite{breiman2001} for performing classification/regression.  
Although decision trees had been previously explored in the literature, 
it was the success of ``boosting''-style machine learning 
techniques, such as AdaBoost \cite{schapire1990,freund1997}, which influenced 
the aggregation of such decision trees into ``forests'' 
with randomized node optimization for improved
classification/regression performance \cite{ho1995,amit1997}.
The final element of bootstrap aggregating or ``bagging'' (i.e.
random sampling of the training data) was
introduced by Breiman \cite{breiman1996} to achieve improved
accuracy.%
\footnote{
One of the principal advantages of R is the extensive community of
developers  who have contributed on the order of thousands of packages 
extending R's capabilities beyond its core functionality.
Most relevant 
is the {\tt randomForest} package developed from Breiman's original
Fortran code by Liaw and Wiener \citep{liaw2002}.
}

@@

%In subsequent sections, we detail the generic
%random forest modeling framework which takes as input
%a set of feature images (also described in a later section)
%and labels for each training subject 
%and outputs a statistical model.  This model, in turn, can then
%be used to provide a probabilistic estimate of the labels in an
%unlabeled subject.

One of the innovations that we provide in this work is
the use of concatenated random forests for improved probabilistic estimation
of the brain segmentation labels.  Briefly, Gaussian mixture
modeling (GMM) is used to create a subset of feature images which are
used to train and predict during the first classification stage.  
The random forest voting output probabilistic maps are then used as 
spatial probability priors for performing a MAP-MRF segmentation
and subsequent feature image creation
thereby replacing the GMM-derived feature images for a second
classification stage.  This dual random forest model generation
and application
improves results and is detailed in later sections.

%As we demonstrate in the Results section, the set of feature
%images employed work sufficiently well for good performance
%on the training data.  However, we discovered
%that further improvements could be gained by using the probabilistic 
%label estimates as input to enhance preprocessing before a 
%second round of feature image generation including 
%modality-specific, prior-based $n$-tissue segmentation and 
%random forest model creation.

@@

Training data consisted of multi-modal brain MRI (T1, T2, FLAIR, and 
post-Gadolinium T1) from 30 glioma patients (both low, $n=10$, and high-grade, $n=20$,
and with and without resection).  For each subject, the T1, T2, and 
FLAIR MRI were linearly registered to the post-contrast T1.  Subsequently,
the brains were skull-stripped and resampled to 1 mm isotropic resolution.
Testing data was processed similarly and released during the course of the
challenge in two sets denoted as ``Leaderboard'' and ``Challenge'' data.  
The former consisted of 21 and 4 high and low-grade tumor patients, respectively,
whereas the latter comprised 10 high-grade only patients.

@@

In terms of implementation, this algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits parallel processing on
an individual workstation or on a large computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the Kirby multivariate symmetrical template consisting of FA, FLAIR,  
T1, and T2 components.  Although DTI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep[e.g.][]{price2003} to warrant investigation in our future work.

@@

\begin{figure*}
  \centering
  \makebox[\textwidth][c]{
  \includegraphics[width=180mm]{Figures/featureImages.pdf}
  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  ``challenge'' data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   
           This mapping is also used to produce the log Jacobian image.  Finally, the 
           (T1 - TC) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}

%For both training and prediction protocols, feature images are generated
%for each subject using the selected input MRI.  
Key to any supervised regression or classification protocol are the 
selected features for training and subsequent prediction.  Based on previous
work and our own experience, the following feature images are generated
for use in both training and prediction:
\begin{itemize}
  \item Per modality (FLAIR, T1, T1C, T2)
    \begin{itemize}
      \item First-order neighborhood statistical images:
            mean, variance, skewness, and entropy. 
            Neighborhood radius $\in \{1,3\}$.
    \item GMM (stage 1) and MAP-MRF (stage 2) posteriors: CSF, gray matter, white 
          matter, necrosis, edema, non-enhancing tumor and enhancing tumor (or a
          subset for the simulated data).
    \item GMM (stage 1) and MAP-MRF (stage 2) connected component geometry 
          features:  distance to tumor core label, volume, volume to surface area ratio, eccentricity, and elongation
    \item Template-based:  symmetric template difference and 
          contralateral difference with Gaussian smoothing ($\sigma = 4$mm).
    \end{itemize}
  \item Miscellaneous: normalized Euclidean distance based on cerebral mask,
    log Jacobian image, and  (T1 - T1C) difference image.
\end{itemize}
The generation of all feature images is performed using the available script 
{\tt createFeatureImages.sh} which also preprocesses the input images performing N4
bias correction \citep{tustison2010} and intensity rescaling to $[0,1]$.  Note that 
it is easily expandable to include 
additional modalities and/or neighborhood and smoothing parameters.
In Figure \ref{fig:featureImages}
we provide sample mid-axial slices from the BRATS\_HG0301
data set from the Challenge cohort.

We previously mentioned that all data was provided already 
skull-stripped.  However, we note to the reader that an
available ANTs-based solution to the brain extraction problem
is found in the script {\tt antsBrainExtraction.sh} which uses
a hybrid approach based on image registration and binary morphology
to isolate the brain \cite{avants2010a}.  We recently performed an evaluation of over 1200
MRI which demonstrated perfect performance \cite{tustison2013a}.

\paragraph{Stage 1 (GMM)}
Prior  cluster centers for specific tissue types learned from training data \citep{reynolds2009} are used in the GMM random forest stage to create 
multiple feature images.  
Given $M$ tissue types, a GMM formulates the 
probability distribution at each voxel, $\mathbf{x}$, as the
sum of Gaussian components, $\mathcal{N}(\mathbf{x}|\mu,\sigma)$, i.e.
\begin{align}
p\left(\mathbf{x}|\mu_m,\sigma_m,\lambda_m\right) = \sum_{i=1}^M \lambda_m \mathcal{N}(\mathbf{x}|\mu_m,\sigma_m)
\end{align}
where $\sum_{m=1}^M \lambda_m = 1$.  One popular method for 
determining the parameters of the GMM is maximum likelihood 
estimation which can performed using the Atropos segmentation 
tool \citep{avants2011} available in ANTs.

\paragraph{Stage 2 (MAP-MRF):}

Prediction during the GMM stage results in a set of random 
forest voting classification maps giving a probabilistic estimate
of the label value at each voxel.  These spatial probability
maps can be used as spatial priors in a MAP-MRF segmentation
estimation protocol also performed using the Atropos segmentation 
tool \citep{avants2011} available in ANTs although we couple
it with iterative bias correction as contained in the script
{\tt antsAtroposN4.sh}.  The probabilistic output (and corresponding
feature images) 
for each modality are used to replace the analogous GMM-based feature images used 
during the first stage of prediction along with the other feature
images previously generated (such as the template-based data).
We discovered that this composite application of random
forest models improves the resulting labeled output.

For both stages, in contrast to previous generative
modeling approaches for multi-modal tumor segmentation 
\citep[e.g.][]{prastawa2003}, we do not use multivariate 
Gaussians to specify tissue probabilities but rather incorporate each
univariate probability map into the feature vector of the training
data.  As pointed out in \cite{menze2010}, multivariate modeling
might obscure the distinct biological information provided by each 
modality.  Instead, we let the random forest construction 
process determine the optimal combination of such multivariate
information.  Additionally, maximum posterior labeling from both stages
is used to determine the connected components for each label.  
Geometric features (assigned voxel-wise) include the physical volumes 
of each connected component, the volume to surface area ratio, 
the elongation, and eccentricity.

\paragraph{Neighborhood First-Order Statistic Feature Images}
Similar to other work \cite{bauer2012,zikic2012} we calculated
feature images using first order statistics such as mean, 
variance, skewness, and entropy.  Images were created for 
neighborhood radii of 1 and 3 voxels.

\paragraph{Template-based Feature Images}
Additional feature images are derived from the multivariate 
symmetric template described earlier. The T1 images of the 
subject and symmetric template are co-registered 
\citep{tustison2013a} using the {\tt antsRegistration} 
tool.  The motivation is similar in spirit to the method
using reflection about the mid-sagittal plane described in 
\cite{geremia2012}.  However, tumor growth often causes 
distortion of the mid-plane thereby limiting the accuracy of
this method.  By mapping to the symmetric template using the
highly accurate SyN algorithm \cite{avants2011a}, we can 
account for any distortion for comparing contralateral
sides.

Two images per modality use this derived mapping.  The modality-specific
contralateral difference feature image is calculated by transforming
the image to the symmetric template, flipping the image contralaterally,
then transforming it back using the inverse transform, and subtracting
the corresponding template component from the original image.  Additionally, 
the template difference image is calculated by subtracting the warped template 
image from the corresponding modality of the subject.

\paragraph{Miscellaneous Feature Images}
Two additional feature images are calculated which seem to aid in 
classification as measured by the random forest variable importance
(relative importance plots are provided in the Results section).
First, the binary mask image is used to calculate the distance image 
\citep{maurer2003}.  Second, from the symmetric template deformation 
mapping we calculate the log Jacobian image.  Related to the earlier
distance feature image, we calculate the distance mask image of the symmetric
template and then warp that distance image to the individual subject.

@@

For use with the Challenge and Leaderboard data, cohort-specific models (both 
low-grade and 
high-grade glioma) for both GMM and MAP-MRF stages were created using only the 
supplied training data.  Initial testing of our proposed framework was performed 
on the training data using a leave-one-out strategy.  Once the
feature images are created for each subject, the resulting images of the entire
training cohort are organized in a csv file for input into the R script
{\tt createModel.R}.  Other possible input parameters include the requested 
number of trees, number of samples per label, and number of threads for parallel
processing.  The output is an {\tt RData} file describing the random forest
model which can be used for future predictions.

Additionally, the {\tt randomForest} package provides a couple measurements 
for determining the importance of chosen features when constructing the model.  
This aids in potential feature pruning or intuiting model behavior.  As mentioned
earlier, we provide the R script {\tt plotVariableImportance.R} to render
one such quantity denoted as {\tt MeanDecreaseAccuracy}.  During model construction
(specifically the out-of-bag error calculation stage), the decrease in prediction accuracy
with the omission of a single feature or variable is tracked and averaged.  Thus,
those features which have the greatest decrease in mean accuracy are considered
to be the most discriminative.%
\footnote{
There is an issue with consistently ranking correlated predictors as described at {\tt http://www.r-bloggers.com/random-forest-variable-importance/} since the permutation testing performed for predictive accuracy assessment assumes predictor independence.  Correctives have been proposed but we ignore these issues for this particular application.
}

\subsubsection{Prediction:  Applying the Random Forest Models}

\begin{figure}
  \centering
  \begin{tabular}{c}
    \includegraphics[width=90mm]{Figures/pipeline.pdf}
  \end{tabular}
  \caption{Diagrammatic workflow for the proposed random forest model prediction.  The feature
  images are first generated from the input MRI and symmetric multivariate template. 
  After organization of the feature images in a csv file, the stage 1 random forest model is 
  applied which results in a first stage classification voting output.  This output
  is used to generate an additional set of feature images for the second classification
  stage.  The final refinement process entails a series of binary morphological 
  operations which results in the final labeled brain.  In the diagram we also list
  the files used to perform each set of functions which are available as open source.
  }
  \label{fig:pipeline}
\end{figure}

Once the models are created, classification of tumors in new subjects is performed
as illustrated in Figure \ref{fig:pipeline}.  From the feature images and input 
GMM model, a tentative set of random forest voting output confidence images are produced.
As described, this is used as input to the second prediction round.  The 
final probability output images are used to produce the maximum labeling.

A final round of binary morphological operations were heuristically designed
to improve the final segmentation results such as removal of small connected 
components and morphological closure of certain regions.
All these steps are included in the perl script 
{\tt applyTumorSegmentationModelForCohort.pl} designed for parallel subject 
processing on the computational cluster at the University of Virginia.%
\footnote{
http://www.uvacse.edu
}

@@

\begin{table*}
\caption{Results from the MICCAI 2013 BRATS Challenge}
\label{table:results}
\begin{center}
\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill} } c c c c c c c c c c c}
\toprule
{\bf Data set} & {\bf Rank} & \multicolumn{3}{c}{\bf Dice} & \multicolumn{3}{c}{ \bf Positive Predictive Value} & \multicolumn{3}{c}{ \bf Sensitivity} \\
{} & {} & complete & core & enhanced & complete & core & enhanced & complete & core & enhanced \\
\midrule
Challenge & 1 & 0.87 (1) & 0.78 (1) & 0.74 (1) & 0.85 (2) & 0.74 (5) & 0.69 (4) & 0.89 (2) & 0.88 (1) & 0.83 (1) \\
Leaderboard & 1 & 0.79 (2) & 0.65 (1) & 0.53 (3) & 0.83 (1) & 0.70 (2) & 0.51 (3) & 0.81 (4) & 0.73 (2) & 0.66 (2) \\
Evaluation & 2 & 0.88 (2) & 0.76 (2) & 0.55 (3) & 0.88 (3) & 0.80 (5) & 0.65 (3) & 0.89 (3) & 0.79 (3) & 0.53 (3) \\
\bottomrule
\end{tabular*}
\end{center}
\end{table*}

A total of four random forest models were created from the 30 training data 
sets. Two models for Stage 1 and Stage 2 processing were generated from the 
20 high-grade glioma evaluation data described earlier.  Similarly, two
additional models were created from the 10 low-grade glioma data sets.  Following
model construction, weighted importance feature plots described earlier were 
rendered to provide feedback as to the individual potential predictive accuracy
of each feature.  The high-grade plots are given in Figure \ref{fig:hgimportance}
whereas the low-grade plots are given in Figure \ref{fig:lgimportance}.

\begin{figure*}
  \makebox[\textwidth][c]{
  \begin{tabular}{cc}
  \includegraphics[width=90mm]{Figures/BRATS_HG_GMM.pdf} &
  \includegraphics[width=90mm]{Figures/BRATS_HG_MAP_MRF.pdf} \\
  \end{tabular}
  }
  \caption{{\tt MeanDecreaseAccuracy} plots generated from the high-grade glioma
  Stage 1 and Stage 2 random forest models.  These plots provide a weighted 
  ranking describing the importance of each feature for predictive accuracy. 
  Features are listed from most to least important.
  }
  \label{fig:hgimportance}
\end{figure*}

\begin{figure*}
  \makebox[\textwidth][c]{
  \begin{tabular}{cc}
  \includegraphics[width=90mm]{Figures/BRATS_LG_GMM.pdf} &
  \includegraphics[width=90mm]{Figures/BRATS_LG_MAP_MRF.pdf} \\
  \end{tabular}
  }
  \caption{{\tt MeanDecreaseAccuracy} plots generated from the low-grade glioma
  Stage 1 and Stage 2 random forest models.  These plots provide a weighted 
  ranking describing the importance of each feature for predictive accuracy.
  Features are listed from most to least important.
  }
  \label{fig:lgimportance}
\end{figure*}

Immediately apparent from these plots are the importance of certain features.
In general, the features based on the symmetric template are quite
discriminative thus justifying the increased computational 
resources required to generate such features.  For single-threaded processing
on the cluster, creating the feature images for a single subject required 
approximately 2 hours of total processing time of which the template registration 
component took approximately 75\%.  Additionally, the first order 
statistical features also seemed fairly important which is something that
previous work had demonstrated \citep[e.g.][]{bauer2012,zikic2012}.
There are a number of differences, however, between low-grade and high-grade 
importance outcomes which would also seem to justify the creation of separate
glioma class models.
